---
title: "Oh Rats"
format:
  pdf:
    latex-auto-install: false
    geometry:
      - top=10mm
      - bottom=20mm
author: Jenna Worthen
execute:
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

# 2026 case study competition

Jenna's first attempt at solving the rats. Help with Grok AI. While
eavesdropping in the TA lab, I noticed TIm had a few of the same plots I
made... **so this definitely** cannot be the final results or plots.

I also want to use more Stat 469 stuff so we'll need to look there in
our meeting.

Extra data sources: [Water
Map](https://www.nyc.gov/content/planning/pages/resources/datasets/borough-boundaries)
\| [NYC City Council Districts
Map](https://hub.arcgis.com/datasets/DCP::nyc-city-council-districts-1/about)
\| [Subway
Stations](https://catalog.data.gov/dataset/mta-subway-stations)

```{r, warning=FALSE, message=FALSE}
library(readxl)
library(spmodel)
library(ggplot2)
library(tidyverse)
library(leaflet)
library(viridis)
library(sf)
library(spdep)
library(patchwork)
library(spatialreg)
library(gstat)
rat <- read.csv("C:/Users/Jenna/OneDrive/Desktop/cc26/2026/311_rodent_calls.csv")
nyc <- read_excel("C:/Users/Jenna/OneDrive/Desktop/cc26/2026/nyc_council_context.xlsx")
```

We are only focusing on NYC so we cannot use this NY state map. Below I
download a shapefile from a arcgis website of nyc council districts.

```{r, warning=FALSE, message=FALSE}
districts <- st_read("C:/Users/Jenna/OneDrive/Desktop/cc26/NYC_City_Council_Districts/NYC_City_Council_Districts.shp")
# ggplot(data = districts) +
#   geom_sf(aes(fill = as.factor(CounDist)), color = "white", size = 0.2) +
#   scale_fill_viridis_d(guide = "none") + # Uses a nice color scale
#   labs(title = "NYC City Council Districts",
#        caption = "Source: NYC Open Data") +
#   theme_minimal()
```

I plot all rat occurenses onto the shapefile colored by district.

```{r}
rats_sf <- rat %>%
  # Remove rows where either Longitude or Latitude is NA
  filter(!is.na(Longitude) & !is.na(Latitude)) %>% 
  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326)

# Transform points to match the NYC Districts shapefile projection
rats_sf <- st_transform(rats_sf, st_crs(districts))

ggplot() +
  # Layer 1: The NYC Districts (Polygons)
  geom_sf(data = districts, fill = "lightgrey", color = "black") +
  # Layer 2: Your Points
  geom_sf(data = rats_sf, aes(color = as.factor(Council.District)), size = 3) +
  theme_minimal()
```

Grok AI said that the best way to begin would be by looking at the
resolution descritions and finding out how many broad categories there
are of resolution. I chose Confirmed and Not confirmed of violations
found/approved rodent problems.

```{r}
description <- rat %>%
  count(`Resolution.Description`, sort = TRUE)
print(description)
```

Here we sort all resolution descriptions into the 2 categories and get a
simple summary.

```{r}
rodent <- rat %>%
  mutate(
    confirmed = case_when(
      str_detect(Resolution.Description, "The Department of Health and Mental Hygiene found violations on the property you reported. Follow-up inspections will be scheduled.") | 
        str_detect(Resolution.Description, "An inspection of this property was conducted and violations were identified. The owner is correcting the condition. If the condition persists, please submit another complaint to 311.")
      # str_detect(`Resolution.Description`, "found violations") |
      # str_detect(`Resolution.Description`, "violations were identified") |
      # str_detect(`Resolution.Description`, "passed with minor violations") |
      # str_detect(`Resolution.Description`, "provided with information and guidance on eliminating rodents) 
      ~ "Confirmed",
      TRUE ~ "Not Confirmed"
    )
  )

# Overall summary (how many are confirmed vs not)
overall_summary <- rodent %>%
  count(confirmed) %>%
  mutate(percent = round(n / sum(n) * 100, 1))

print("=== Overall Summary ===")
print(overall_summary)

# District-level summary (this is the important one)
district_summary <- rodent %>%
  count(`Council.District`, confirmed) %>%          # make sure the column name matches exactly
  pivot_wider(names_from = confirmed, 
              values_from = n, 
              values_fill = 0) %>%
  mutate(
    total_reports = Confirmed + `Not Confirmed`,
    confirmed_issues = Confirmed
  ) %>%
  select(`Council.District`, total_reports, confirmed_issues) %>%
  arrange(desc(total_reports))

print("=== District Summary (top 10 shown) ===")
print(district_summary %>% slice_head(n = 10))
print("Full district_summary has 51 rows")
```

Now we separate these statistics for all districts. We find the total
rat reports and total confirmed rat reports.

```{r}
# 2. Aggregate to district level
district_rodent <- rodent %>%
  count(`Council.District`, confirmed) %>%
  pivot_wider(names_from = confirmed, values_from = n, values_fill = 0) %>%
  mutate(
    total_reports = Confirmed + `Not Confirmed`,
    confirmed_issues = Confirmed
  ) %>%
  select(district_id = `Council.District`, 
         total_reports, 
         confirmed_issues)

# 4. Join them (make sure the ID column names match)
combined <- nyc %>%
  left_join(district_rodent, by = c("council_district_id" = "district_id")) %>%   # <-- CHANGE THIS LINE IF THE COLUMN NAME IS DIFFERENT
  mutate(
    population = as.numeric(population),   # just in case
    rate_all_reports_per1000 = (total_reports / population) * 1000,
    rate_confirmed_per1000   = (confirmed_issues / population) * 1000,
    rate_all_reports = (total_reports / population),
    rate_confirmed   = (confirmed_issues / total_reports),
    density = (housing_units / population)
  ) %>%
  arrange(desc(rate_all_reports_per1000))

# 5. Print key results
print("=== Top 10 districts by reporting rate (per 1,000 people) ===")
combined %>%
  select(council_district_id, council_district_name, population, 
         total_reports, rate_all_reports_per1000, 
         confirmed_issues, rate_confirmed_per1000) %>%
  slice_head(n = 10) %>%
  print()

print("\n=== Summary statistics of the two rates ===")
summary(combined %>% select(rate_all_reports_per1000, rate_confirmed_per1000))
```

```{r}
map_data <- districts %>%
  mutate(CounDist = as.numeric(CounDist)) %>%   # ensure it's numeric
  left_join(combined, by = c("CounDist" = "council_district_id"))

# 4. Side-by-side maps
p1 <- ggplot(map_data) +
  geom_sf(aes(fill = rate_all_reports_per1000), color = "white", linewidth = 0.3) +
    geom_sf_text(aes(label = CounDist), size = 3, color = "red3", fontface = "bold") +
  scale_fill_viridis_c(option = "plasma", name = "Reports per 1,000 residents") +
  labs(title = "All Rodent 311 Reporting Rate",
       subtitle = "2025 • Hotter colors = more complaints per capita") +
  theme_minimal() +
  theme(legend.position = "bottom", plot.title = element_text(face = "bold"))

p2 <- ggplot(map_data) +
  geom_sf(aes(fill = rate_confirmed), color = "white", linewidth = 0.3) +
    geom_sf_text(aes(label = CounDist), size = 3, color = "red3", fontface = "bold") +
  scale_fill_viridis_c(option = "turbo", name = "Confirmed Issues / Total") +
  labs(title = "Confirmed Rodent Issues",
       subtitle = "Violations found or owner guidance provided (27% of reports)") +
  theme_minimal() +
  theme(legend.position = "bottom", plot.title = element_text(face = "bold"))

combined_map <- p1 + p2 + plot_layout(ncol = 2)
print(combined_map)
```

## GGPlot 2 Statistics

Here is a patchwork plot with the 2 statistics for each district.

```{r}
map_data <- districts %>%
  mutate(CounDist = as.numeric(CounDist)) %>%   # ensure it's numeric
  left_join(combined, by = c("CounDist" = "council_district_id"))

# 4. Side-by-side maps
p1 <- ggplot(map_data) +
  geom_sf(aes(fill = rate_all_reports_per1000), color = "white", linewidth = 0.3) +
    geom_sf_text(aes(label = CounDist), size = 3, color = "red3", fontface = "bold") +
  scale_fill_viridis_c(option = "plasma", name = "Reports per 1,000 residents") +
  labs(title = "All Rodent 311 Reporting Rate",
       subtitle = "2025 • Hotter colors = more complaints per capita") +
  theme_minimal() +
  theme(legend.position = "bottom", plot.title = element_text(face = "bold"))

p2 <- ggplot(map_data) +
  geom_sf(aes(fill = rate_confirmed_per1000), color = "white", linewidth = 0.3) +
    geom_sf_text(aes(label = CounDist), size = 3, color = "red3", fontface = "bold") +
  scale_fill_viridis_c(option = "inferno", name = "Confirmed Issues per 1,000 residents") +
  labs(title = "Confirmed Rodent Issues Rate",
       subtitle = "Violations found or owner guidance provided (27% of reports)") +
  theme_minimal() +
  theme(legend.position = "bottom", plot.title = element_text(face = "bold"))

combined_map <- p1 + p2 + plot_layout(ncol = 2)
print(combined_map)
```

Same thing as above but with leaflet.

```{r, include = FALSE}
# map_data <- districts %>%
#   mutate(CounDist = as.numeric(CounDist)) %>%
#   left_join(combined, by = c("CounDist" = "council_district_id")) %>%
#   # Transform to lat/long for Leaflet
#   st_transform(4326) %>%
#   # Create nice popup
#   mutate(
#     popup_text = paste0(
#       "<b>District ", CounDist, "</b><br>",
#       "<i>", council_district_name, "</i><br><br>",
#       "All Reports: <b>", total_reports, "</b> (", rate_all_reports_per1000, " per 1,000)<br>",
#       "Confirmed Issues: <b>", confirmed_issues, "</b> (", rate_confirmed_per1000, " per 1,000)<br>",
#       "Population: ", scales::comma(population)
#     )
#   )
# 
# pal_reports <- colorNumeric("plasma", domain = map_data$rate_all_reports_per1000, na.color = "grey")
# pal_conf    <- colorNumeric("inferno", domain = map_data$rate_confirmed_per1000, na.color = "grey")
# 
# leaflet_map <- leaflet(map_data) %>%
#   addProviderTiles(providers$CartoDB.Positron) %>%
#   
#   # Layer 1: All Reporting Rate
#   addPolygons(
#     group = "All Reporting Rate",
#     fillColor = ~pal_reports(rate_all_reports_per1000),
#     weight = 1.5,
#     opacity = 1,
#     color = "white",
#     fillOpacity = 0.75,
#     popup = ~popup_text,
#     highlightOptions = highlightOptions(weight = 3, color = "black", bringToFront = TRUE)
#   ) %>%
#   
#   # Layer 2: Confirmed Rate
#   addPolygons(
#     group = "Confirmed Issues Rate",
#     fillColor = ~pal_conf(rate_confirmed_per1000),
#     weight = 1.5,
#     opacity = 1,
#     color = "white",
#     fillOpacity = 0.75,
#     popup = ~popup_text,
#     highlightOptions = highlightOptions(weight = 3, color = "black", bringToFront = TRUE)
#   ) %>%
#   
#   # Layer controls
#   addLayersControl(
#     baseGroups = c("All Reporting Rate", "Confirmed Issues Rate"),
#     options = layersControlOptions(collapsed = FALSE)
#   ) %>%
#   
#   # Legends
#   addLegend("bottomright",
#             pal = pal_reports,
#             values = ~rate_all_reports_per1000,
#             title = "Reports per 1,000",
#             opacity = 0.9) %>%
#   addLegend("bottomleft",
#             pal = pal_conf,
#             values = ~rate_confirmed_per1000,
#             title = "Confirmed per 1,000",
#             opacity = 0.9)
# 
# leaflet_map   # ← this should now open cleanly in the Viewer
```

```{r}
# 5. Quick correlation table (careful interpretation only!)
cor_table <- combined %>%
  select(rate_all_reports_per1000, rate_confirmed_per1000,
         poverty_rate, unemployment_rate,
         pct_renter_occupied_among_occupied_units, 
         pct_vacant_housing_units, 
         median_household_income,
         pct_households_broadband,
         pct_language_other_than_english_proxy) %>%
  cor(use = "complete.obs") %>%
  round(3)

print("=== Pearson Correlations with Rates ===")
print(cor_table)
```

```{r}
library(spdep)

# Use the combined object from before
model_data <- combined %>%
  filter(!is.na(rate_all_reports_per1000))   # drop any missing if any

# Create spatial weights (queen contiguity)
nb <- poly2nb(map_data, queen = TRUE)          # map_data from Leaflet code (has geometry)
lw <- nb2listw(nb, style = "W", zero.policy = TRUE)

# ----------------- Model 1: All Reports Rate -----------------
mod_reports <- lm(rate_all_reports_per1000 ~
                    density + 
                    pct_renter_occupied_among_occupied_units +
                    poverty_rate +
                    unemployment_rate +
                    median_household_income+ pct_language_other_than_english_proxy,
                  data = model_data)

summary(mod_reports)

# # Moran’s I on residuals
# moran_reports <- moran.test(residuals(mod_reports), lw, zero.policy = TRUE)
# print("Moran's I for Reports model residuals:")
# print(moran_reports)

# ----------------- Model 2: Confirmed Issues Rate -----------------
mod_confirmed <- lm(rate_confirmed_per1000 ~ 
                      density + 
                      pct_renter_occupied_among_occupied_units +
                      poverty_rate +
                      unemployment_rate +
                      median_household_income + pct_language_other_than_english_proxy,
                    data = model_data)

summary(mod_confirmed)

# moran_conf <- moran.test(residuals(mod_confirmed), lw, zero.policy = TRUE)
# print("Moran's I for Confirmed model residuals:")
# print(moran_conf)
```

**Reporting Rate Model (All 311 complaints per 1,000)**

\% Renter-occupied is the only statistically significant predictor (p =
0.0035). For every 10 percentage point increase in renter-occupied
units, the reporting rate increases by about 0.68 per 1,000 residents.
Poverty, unemployment, and median income are not significant once renter
% is in the model. Model explains \~40% of the variation (Adjusted R² =
0.35). Strong spatial autocorrelation in residuals (Moran’s I = 0.275, p
= 0.0006) → we should account for spatial dependence.

**Confirmed Issues Rate Model**

\% Renter-occupied is even stronger (p \< 0.0001). Every 10 pp increase
→ +0.27 confirmed issues per 1,000. Other variables remain
non-significant. Model explains \~47% of variation (Adjusted R² = 0.43).
No significant spatial autocorrelation in residuals (Moran’s I p =
0.156) — ordinary least squares is acceptable here.

### Key takeaway (careful framing for DOHMH):

Higher renter-occupied housing share is consistently associated with
both more rodent complaints and more confirmed issues, even after
controlling for poverty, unemployment, and income. This points toward
housing stock and rental market conditions (older buildings, maintenance
issues, tenant turnover, etc.) rather than demographics alone.

## Simple spatial model

```{r}
# Spatial Lag Model (reports rate)
sem_confirmed_reports <- errorsarlm(rate_confirmed_per1000 ~ 
                      density + 
                      pct_renter_occupied_among_occupied_units +
                      poverty_rate +
                      unemployment_rate +
                      median_household_income + pct_language_other_than_english_proxy,
                       data = model_data, 
                       listw = lw, 
                       zero.policy = TRUE)

summary(sem_confirmed_reports)

# Spatial Error Model (reports rate)
sem_reports <- errorsarlm(rate_all_reports_per1000 ~ 
                      density + 
                      pct_renter_occupied_among_occupied_units +
                      poverty_rate +
                      unemployment_rate +
                      median_household_income + pct_language_other_than_english_proxy,
                          data = model_data, 
                          listw = lw, 
                          zero.policy = TRUE)

summary(sem_reports)
```

### Conclusions from the best model (SEM):

After accounting for unobserved spatial clustering (e.g., neighboring
districts sharing similar building ages, sanitation patterns, or
reporting norms), % renter-occupied housing remains the strongest and
only significant predictor. A 10 percentage-point higher renter share is
associated with +0.75 rodent reports per 1,000 residents (and similarly
strong for confirmed issues). Poverty, unemployment, and income drop to
near-zero and non-significant once we control for renter % and spatial
dependence.

**Housing tenure (rental vs owner-occupied stock) appears to be the
dominant district-level driver of both rodent reporting and confirmed
issues in 2025.**

## Validations

```{r}
# Confirmed
map_data$model_resids <- residuals(sem_confirmed_reports)
vgm_resids_c <- variogram(model_resids ~ 1, data = map_data)
plot(vgm_resids_c, main = "Confirmed Rep Variogram")

# Total reports
map_data$model_resids <- residuals(sem_reports)
vgm_resids <- variogram(model_resids ~ 1, data = map_data)
plot(vgm_resids, main = "Total Rep Variogram")
```

```{r}
hist(residuals(sem_confirmed_reports))
hist(residuals(sem_reports))
```

## Predictions vs Actual

```{r}
# # Add predictions and residuals from the best model (SEM)
# model_data <- model_data %>%
#   mutate(
#     pred_sem = fitted(sem_reports),
#     resid_sem = residuals(sem_reports)
#   )
# 
# # Join back to map_data for plotting
# map_final <- map_data %>%
#   left_join(model_data %>% select(council_district_id, pred_sem, resid_sem),
#             by = c("CounDist" = "council_district_id"))
# 
# # 1. Observed vs Predicted side-by-side
# p_obs <- ggplot(map_final) +
#   geom_sf(aes(fill = rate_all_reports_per1000), color = "white") +
#   scale_fill_viridis_c(option = "plasma", name = "Reports per 1,000") +
#   labs(title = "Observed Reporting Rate") +
#   theme_minimal()
# 
# p_pred <- ggplot(map_final) +
#   geom_sf(aes(fill = pred_sem), color = "white") +
#   scale_fill_viridis_c(option = "plasma", name = "Predicted") +
#   labs(title = "Predicted from Spatial Error Model") +
#   theme_minimal()
# 
# p_obs + p_pred
# 
# # 2. Residual map (red = model under-predicts, blue = over-predicts)
# ggplot(map_final) +
#   geom_sf(aes(fill = resid_sem), color = "white") +
#   scale_fill_gradient2(low = "blue", mid = "white", high = "red",
#                        midpoint = 0, name = "Residual") +
#   labs(title = "Residuals from Spatial Error Model",
#        subtitle = "Red = more reports than expected given renter % and spatial effects") +
#   theme_minimal()
# 
# # 3. Print key coefficients for slides
# summary(sem_reports)
```

## New covariates: Adding Distance to water and Subway relevance

```{r}
# 1. Your existing objects (map_data or combined)
# Assume you have map_data from before with geometry

# 2. Load subway stations
subway <- read_csv("C:/Users/Jenna/OneDrive/Desktop/cc26/MTA_Subway_Stations.csv") %>%
  filter(!is.na(Latitude), !is.na(Longitude)) %>%
  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326) %>%
  st_transform(st_crs(map_data))   # match your districts CRS

# 3. Count subway stations per district (intersection)
districts_with_subway <- map_data %>%
  st_join(subway, join = st_intersects) %>%
  group_by(CounDist, council_district_name) %>%
  summarise(
    subway_count = n(),
    .groups = "drop"
  ) %>%
  st_drop_geometry()

# Alternative: distance from district centroid to nearest subway
centroids <- st_centroid(map_data)

dist_to_subway <- st_distance(centroids, subway) %>%
  apply(1, min)   # minimum distance in meters to any station

map_data <- map_data %>%
  mutate(
    dist_to_nearest_subway_m = as.numeric(dist_to_subway),
    subway_density = districts_with_subway$subway_count / as.numeric(st_area(geometry) / 1e6)   # per km² if you add count
  )

# Adding distance to water shorline
# 2. Load the shoreline layer
shoreline <- st_read("C:/Users/Jenna/OneDrive/Desktop/cc26/nybb_26a/nybb.shp") %>%   # or .geojson
  st_transform(st_crs(districts))   # match CRS

# 3. Create district centroids
centroids <- st_centroid(districts)

# 4. Calculate minimum distance to shoreline (in meters)
dist_matrix <- st_distance(centroids, shoreline)
min_dist_to_shore <- apply(dist_matrix, 1, min)

# 5. Add to your map_data / combined
map_data <- map_data %>%
  mutate(dist_to_shoreline_m = as.numeric(min_dist_to_shore))

# Optional: create a categorical version for easier interpretation
map_data <- map_data %>%
  mutate(near_water = ifelse(dist_to_shoreline_m < 500, "Within 500m of shoreline", "Farther inland"))

# Quick check
summary(map_data$dist_to_shoreline_m)
```

```{r}
# 1. Make sure model_data has the same districts as before
model_data <- combined %>%                     # 'combined' is your original joined dataset with rates
  filter(!is.na(rate_all_reports_per1000))

# 2. Add the new water variable from map_data
#    (map_data has the geometry + CounDist)
water_vars <- map_data %>%
  st_drop_geometry() %>%
  select(CounDist, dist_to_shoreline_m) %>%    # add any other water vars if you created them
  rename(council_district_id = CounDist)

# 3. Join into model_data
model_data <- model_data %>%
  left_join(water_vars, by = "council_district_id")

# 4. Quick sanity check
summary(model_data$dist_to_shoreline_m)
head(model_data %>% select(council_district_id, rate_all_reports_per1000, 
                           pct_renter_occupied_among_occupied_units, 
                           dist_to_shoreline_m))

# 5. Re-run the best Spatial Error Model with the new variable
sem_water <- errorsarlm(rate_all_reports_per1000 ~ 
                          pct_renter_occupied_among_occupied_units + 
                          dist_to_shoreline_m +
                          poverty_rate + 
                          unemployment_rate + 
                          median_household_income,
                        data = model_data, 
                        listw = lw, 
                        zero.policy = TRUE)

summary(sem_water)

# Optional: also try a version with confirmed rate
sem_conf_water <- errorsarlm(rate_confirmed_per1000 ~ 
                               pct_renter_occupied_among_occupied_units + 
                               dist_to_shoreline_m +
                               poverty_rate,
                             data = model_data, 
                             listw = lw, 
                             zero.policy = TRUE)

summary(sem_conf_water)
```

```{r, warning = FALSE, message = FALSE}
# ====================== 1. Subway stations ======================
subway <- read_csv("C:/Users/Jenna/OneDrive/Desktop/cc26/MTA_Subway_Stations.csv") %>%
  filter(!is.na(Latitude), !is.na(Longitude)) %>%
  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326) %>%
  st_transform(st_crs(map_data))

# Count stations per district (intersection)
districts_with_subway <- map_data %>%
  st_join(subway, join = st_intersects) %>%
  group_by(CounDist, council_district_name) %>%
  summarise(subway_count = n(), .groups = "drop") %>%
  st_drop_geometry()

# Distance to nearest subway from district centroid
centroids <- st_centroid(map_data)
dist_to_subway <- st_distance(centroids, subway) %>%
  apply(1, min)

# ====================== 2. Shoreline (your file) ======================
shoreline <- st_read("C:/Users/Jenna/OneDrive/Desktop/cc26/nybb_26a/nybb.shp") %>%
  st_transform(st_crs(map_data))

dist_matrix <- st_distance(centroids, shoreline)
min_dist_to_shore <- apply(dist_matrix, 1, min)

# ====================== 3. Add all new variables to map_data ======================
map_data <- map_data %>%
  mutate(
    subway_count = districts_with_subway$subway_count,
    dist_to_nearest_subway_m = as.numeric(dist_to_subway),
    dist_to_shoreline_m = as.numeric(min_dist_to_shore),
    near_water = ifelse(dist_to_shoreline_m < 500, "Within 500m", "Farther inland")
  )

# Quick summary of new variables
summary(map_data %>% select(subway_count, dist_to_nearest_subway_m, dist_to_shoreline_m))

# ====================== 4. Update model_data ======================
model_data <- combined %>%
  filter(!is.na(rate_all_reports_per1000)) %>%
  left_join(
    map_data %>% 
      st_drop_geometry() %>%
      select(CounDist, subway_count, dist_to_nearest_subway_m, 
             dist_to_shoreline_m, near_water),
    by = c("council_district_id" = "CounDist")
  )

# ====================== 5. Re-run Spatial Error Model with new variables ======================
sem_full <- errorsarlm(rate_all_reports_per1000 ~ 
                         pct_renter_occupied_among_occupied_units +
                         subway_count +
                         dist_to_shoreline_m +
                         poverty_rate +
                         unemployment_rate +
                         median_household_income,
                       data = model_data, 
                       listw = lw, 
                       zero.policy = TRUE)

summary(sem_full)

# Also for confirmed rate (optional but useful)
sem_conf_full <- errorsarlm(rate_confirmed_per1000 ~ 
                              pct_renter_occupied_among_occupied_units +
                              subway_count +
                              dist_to_shoreline_m +
                              poverty_rate,
                            data = model_data, 
                            listw = lw, 
                            zero.policy = TRUE)

summary(sem_conf_full)
```

### Key Takeaways from the Full Model

**For All Reports Rate (best Spatial Error Model)":**

\% Renter-occupied remains very strong and highly significant (coeff ≈
7.28, p \< 0.00001). → Every 10 percentage point increase in renter
share → roughly +0.73 reports per 1,000 residents. Subway count and
distance to shoreline are not statistically significant (p = 0.139 and p
= 0.451). Lambda is still strong and significant → spatial clustering
matters. AIC is similar to the previous model without subway/water
(200.21 vs 199.00), so the new variables don’t improve the model much.

**For Confirmed Issues Rate:**

Again, only % renter-occupied is strongly significant. Subway count and
distance to shoreline are clearly not significant. Poverty is borderline
(p ≈ 0.053), but weak.

**Bottom line:** The extra features (subway stations and proximity to
water) do not add meaningful explanatory power once we account for
renter housing share and spatial dependence. This is actually useful
information — it suggests that in 2025 NYC rodent
complaints/confirmations are more strongly tied to housing tenure and
rental stock characteristics than to subway density or shoreline
proximity at the council-district level.
