---
title: "Linguistics EDA R Answers"
format: pdf
editor: visual
execute:
  message: false
  echo: false
---

```{r, echo=FALSE, message=FALSE}
library(tidyverse)
library(vroom)
library(spmodel)
library(viridis)
dat <- vroom('../Data/All_Linguistics.csv') %>%
  filter(!is.na(y))
```

1.  Create a heat map of the linguistics data. Use ChatGPT to figure out how to add a map of the US to the background. Also, for the color scale, use one of the `viridis` color scales (which are specially tailored to people with color blindness).

```{r}
us_states <- map_data("state")
ggplot() + 
  geom_polygon(data = us_states, aes(x = long, y = lat, group = group), 
               fill = NA, color = "black") +
  geom_point(data=dat, aes(x=long, y=lat, color=y)) +
  scale_color_viridis()
```

2.  Create and plot the auto-*covariance* function of residuals from a linear model that uses `state` to predict the Getis-ord score (`y`) with the linguistics data.

```{r}
dat %>%
  eacf(formula=y~state, data=., xcoord=long, ycoord=lat) %>%
  plot()
```

3.  To show what a auto-covariance function looks like for data with NO spatial correlation, shuffle the longitude and latitude coordinates of the linguistics data and recompute and plot the resulting `eacf` from #2 above.

```{r}
shuffled_dat <- dat %>%
  mutate(lat=sample(lat),
         long=sample(long))
shuffled_esv <- eacf(formula=y~state, data=shuffled_dat, xcoord=long, ycoord=lat)
plot(shuffled_esv)

```
